---
title: "Prepare Gene Sets for Enrichment Analysis"
author: "Tyler Sagendorf"
date: "Last compiled `r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Prepare Gene Sets for Enrichment Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 4,
  fig.width = 6
)
```

```{r setup}
# Required packages
library(ostRich)
```


# Create List of Gene Sets from GMT File

```{r read-gmt}
# Path to GMT file - MSigDB Gene Ontology sets
gmt_file <- system.file("extdata", "c5.go.v2023.2.Hs.symbols.gmt.gz",
                        package = "ostRich")

gene_sets <- gmt_to_list(gmt_file)

class(gene_sets)

length(gene_sets) # 10461 gene sets

head(names(gene_sets))

gene_sets[1] # genes in first set
```

# Filter Gene Sets

We will restrict gene sets to only those genes in `background` (usually the genes from a particular experiment). Only those sets with at least 10 genes that do not contain all background genes will be retained.

Normally, the background would be the vector of unique genes that were detected in a particular -omics experiment. For example purposes, we will set the background to the top 4000 most common genes. This ensures that at least some gene sets will have a high degree of overlap, which is necessary for clustering later.

```{r}
# Top 4000 most common genes
top_genes <- table(unlist(gene_sets))
top_genes <- sort(top_genes, decreasing = TRUE)[1:4000]
head(top_genes)

background <- names(top_genes)
```

```{r filter-sets}
# Filter sets
gene_sets_filt <- filter_sets(gene_sets, 
                              background = background, 
                              min_size = 10L,
                              max_size = length(background) - 1L)
length(gene_sets_filt) # 5701 (down from 10461)
```

We can plot the ratio of the new set sizes to their original sizes. This ratio serves as a measure of confidence that what genes remain for testing is correctly described by the gene set name.

```{r}
sizes_new <- lengths(gene_sets_filt)
sizes_old <- lengths(gene_sets)[names(sizes_new)]
r <- sizes_new / sizes_old

hist(r, breaks = seq(0, 1, 0.05))
```

We can determine what minimum proportion of genes are retained by the top 90% of sets.

```{r}
quantile(r, 1 - 0.9) # 0.3870968
```

The top 90% of gene sets retain at least `r round(quantile(r, 1 - 0.9) * 100, 1)`% of their original genes.


# Cluster Gene Sets

Now, we will cluster highly similar sets. This helps to reduce redundancy of enrichment analysis/gene set testing results. At the very least, this should be used to remove aliased sets by setting `cutoff` to 1. Two or more sets are aliased if they contain the same genes, but they have different descriptions.

The `cutoff` is the minimum Jaccard similarity coefficient (default `method`) used to classify two sets as sufficiently similar. A larger `h` (cut height) will increase the maximum size of clusters. In practice, we use the default 0.9.

```{r cluster-sets}
# Cluster sets with default parameters. Same as cluster_sets(gene_sets_filt)
cluster_df <- cluster_sets(gene_sets_filt, 
                           method = "jaccard", 
                           cutoff = 0.85,
                           h = 0.9)
```

```{r}
# Include size_ratio and n_char columns
cluster_df$size_ratio <- r[cluster_df$set]
cluster_df$n_char <- nchar(cluster_df$set)

# Reorder rows
o <- with(cluster_df, order(cluster, -set_size, -size_ratio, n_char, set))
cluster_df <- cluster_df[o, ]
```

Due to how `cluster_sets` works, clusters with at least 2 gene sets will always appear first. Therefore, if we select the data for the first 5 clusters, this should show groups of highly similar sets (unless no pairs of sets were sufficiently similar, according to the choice of `cutoff`).

```{r}
# Show data for first 5 clusters
cluster_df[cluster_df$cluster <= 5, ]
```

```{r}
# Total number of clusters
max(cluster_df$cluster) # 5478

# Number of clusters with n sets
barplot(table(table(cluster_df$cluster)),
        xlab = "Number of Sets per Cluster",
        ylab = "Number of Clusters")
```

From the bar plot, most clusters consist of singleton sets. The largest cluster(s) contain 4 sets.

Now, we will keep the first set from each cluster. To do this, we can remove any sets where the cluster is duplicated.

```{r}
# Select first set from each cluster
keep_sets <- cluster_df$set[!duplicated(cluster_df$cluster)]
head(keep_sets)

# Filter sets once more
gene_sets_filt <- gene_sets_filt[keep_sets]
length(gene_sets_filt) # 5478 (down from 5701)
```


Now, `gene_sets_filt` can be used with any enrichment analysis/gene set testing procedure of your choice. We recommend `camera` or `cameraPR` from the limma package. CAMERA is an extension of the t-test that accounts for inter-gene correlation to correctly control the type I error rate.


# Session Info

```{r}
sI <- sessionInfo()
print(sI, locale = FALSE)
```

